
## Operators

= / <- 

|> 

# Data Types

Vectors and Lists
### Vectors: Atomic vectors

Characters
Integers
Boolean
Decimals
Factors

For vectors they have to be of the same type
Vectors can be empty or of any length (i.e., 1, 2, ...., n)

For vectors with more than elements/object, you combine using `c()` (combine function) and separate the elements using commas


### Lists
List is also vector
But they can be of different types

### Data frames

Just lists, that are two dimensional
The lists of the same length
Elements of one list should be of the same type


# Lab 1

## Atomic Vectors - Characters

These are objects with strings of any length
For example, names of people, names of places, names of countries
They should be enclosed in quotation marks
R accepts both single and double quotation marks

### length 1 character vectors
```{r characters}
# examples: The following are the same
# with double quotes
first_name <- "john"
print(first_name)
first_name
(first_name <- "john")
# with single quotes
first_name <- 'john'
first_name
```


### length more than 

```{r characters-more-than-one}
# examples
first_names <- c('john', 'cyrus')
first_names


first_names1 <- c('john', 'cyrus', 'josh', 'lucky')
first_names1
```
When you have vectors you can check the number of items using `length()`

#### Checking the legth of our items


```{r character-length}
# length one character vector
length1 <- length(first_name)
# use paste()/paste0() to combine elements when printing
# the difference is how they separate the elements they are combining
paste("Vector with length: ", length1) # will always separate items by space (the default case)
paste("Vector with length: ", length1, sep = ":") # separate with a new line
paste0("Vector with length: ", length1) # prints as provided
paste0("Vector with length:", length1) # prints as provided

# length two character vectors
length(first_name)

# other character vectors
length(first_names1)

```
### Empty Vectors

Empty have no type unless you assign them explicitly

```{r empty}
# option 1
null_name_opt1 <- NULL
null_name_opt1

# option 2
# initialize with nothing
null_name_opt2 <- character(0)
null_name_opt2

length(null_name_opt1)
length(null_name_opt2)
```

### Type checking

You can check the type of the vectors

This can be done using `is.character()` for characters, for integers `is.integer()` and so forth - These returns `TRUE`/`FALSE`

You can also check they type using `typeof()`, this returns where it's character, integer, and so forth

```{r characters}
# we use the characters we defined previous

# using is.character()
is.character(first_name)

# using typeof()
typeof(first_name)



# using is.character()
is.character(first_names1)

# using typeof()
typeof(first_names1)


# using is.character()
is.character(null_name)
typeof(null_name)

# using typeof()
is.character(null_name)
typeof(null_name_opt2)

```

## Atomic Vectors - Numeric

Under numeric vectors we have integers, double, 

Integers finite (countable) numbers which are whole

Double is any number inclusive of integers, and with numbers on the right side of the decimal

```{r}
# only one element
year_integer <- 2020
year_integer
# two  elements
years_integer <- c(2020, 2021)
years_integer
# more than two  elements
years_integer1 <- c(2020, 2021, 2023, 2024)
years_integer1
```

#### Type checking

```{r type-check-integers}
typeof(year_integer)
typeof(years_integer)
typeof(years_integer1)


is.double(year_integer)
is.integer(year_integer)


integer_initialize <- integer(0)
is.integer(integer_initialize)
```

## Factors

In Stata they are the labelled values

Factors are categorical variables with distinct number of items
These includes things like level of education, gender, liker scale items, means of transport work, rank of employees, type of school

Factors are of two types:
  - ordered factors: The order is of importance. Essential they are ranked from lower to higher rank, e.g. level of education, rank of employees
  - unordered factors, the order of items do not matter, for example, marital status, type of school, means of transport work, gender, and so forth
  
You create factors using `factor()`   function
  
### Lab 2

#### Unorder factors

factors with characters

```{r}
factor_gender <- factor(x = c("male", "female"), levels = c("male", "female"))
factor_gender

factor_gender_with_labels <- factor(x = c("male", "female"), levels = c("male", "female"), labels = c("Man", "Woman"))
factor_gender_with_labels


```

Factors with integers

```{r}
factor_gender_int <- factor(x = c(1, 2), levels = c(1, 2))
factor_gender_int

factor_gender_with_labels1 <- factor(x = c(1, 2), levels = c(1, 2), labels = c("Man", "Woman"))
factor_gender_with_labels1
```

### Type chacking -factors

```{r}
typeof(factor_gender)
typeof(factor_gender_int)


is.factor(factor_gender)
```

